\chapter*{Acknowledgements}
\addcontentsline{toc}{chapter}{\protect\numberline{}Acknowledgements}

\noindent This book has been an open-source project since day one. Instructors and students can download the PDF for free and modify the source as they see fit. Several of instructors and students have provided extremely useful feedback, which has improved the book at each iteration. Moreover, due to the open-source nature of the book, I have been able to incorporate content written by others. Below is a partial list of people (alphabetical by last name) that have contributed content, advice, or feedback.

\begin{itemize}
\item \href{https://thalestriangles.blogspot.com}{Joshua Bowman} (Pepperdine University). The Preface is a modified version of Joshua's \emph{An IBL preface}.
\item \href{https://faculty.bentley.edu/details.asp?uname=ncarter}{Nathan Carter} (Bentley University). Nathan's excellent book \emph{Visual Group Theory} has had a huge impact on my approach to teaching abstract algebra.
\item \href{http://home.snc.edu/andershendrickson/}{Anders Hendrickson} (St.~Norbert College). Anders is the original author of the content in Appendix~\ref{appendix:elements_of_style}: Elements of Style for Proofs. The current version in Appendix~\ref{appendix:elements_of_style} is a result of modifications made by myself with some suggestions from Dave Richeson.
\item \href{https://embracinglifewithmajorrevisions.org/2017/07/12/rights-of-the-learner-an-introduction/}{Crystal Kalinec-Craig} (University of Texas at San Antonio). Section~\ref{sec:Rights of the Learner}: Rights of the Learner is an adaptation of a similar list written by Crystal.
\item \href{https://ericwmiles.weebly.com}{Eric Miles} (Colorado Mesa University). Eric modified an earlier version of the book. I reincorporated several of his improvements.
\item \href{http://users.dickinson.edu/~richesod/}{Dave Richeson} (Dickinson College). Dave is responsible for much of the content in Appendix~\ref{appendix:fancy_math_terms}: Fancy Mathematical Terms and Appendix~\ref{appendix:definitions}: Definitions in Mathematics. 
\item \href{http://webpages.csus.edu/wiscons/}{Josh Wiscons} (CSU Sacramento) and \href{http://emp.byui.edu/woodruffb/}{Ben Woodruff} (BYU Idaho). In the early stages of development, Josh and Ben were instrumental the development of these notes.
\end{itemize}
