\chapter{Subgroups and Isomorphisms}
\label{chapter:subgroups_isomorphisms}
\thispagestyle{empty}

For the next two sections, it would be useful to have all of the Cayley diagrams we've encountered in one place for reference. So, before continuing, gather up the following Cayley diagrams:
\begin{itemize}
\item $\Spin_{1\times 2}$. There are 3 of these.  I drew one for you in Section~\ref{sec:cayley_diagrams} and you discovered two more in Problem~\ref{cayley:altSpin1x2}.
\item $S_2$.  See Problem~\ref{prob:make_Cayley_diagrams}\ref{cayley:S2}.
\item $R_4$.  See Problem~\ref{prob:make_Cayley_diagrams}\ref{cayley:R4}.
\item $V_4$.  See Problem~\ref{prob:make_Cayley_diagrams}\ref{cayley:V4}.
\item $D_3$.  There are two of these.  See Problems~\ref{prob:make_Cayley_diagrams}\ref{cayley:D3} and \ref{prob:make_Cayley_diagrams}\ref{cayley:altD3}.
\item $S_3$.  See Problem~\ref{prob:make_Cayley_diagrams}\ref{cayley:S3}.
\item $D_4$.  See Problem~\ref{prob:make_Cayley_diagrams}\ref{cayley:D4}.
\end{itemize}

%%----------------------------%%

\begin{section}{Subgroups}

%%----------------------------%%

\begin{problem}
Recall the definition of ``subset."  What do you think ``subgroup" means?  Try to come up with a potential definition.  Try not to read any further before doing this.
\end{problem}

\begin{problem}\label{prob:R4_in_D4}
Examine your Cayley diagrams for $D_4$ (with generating set $\{r,s\}$) and $R_4$ (with generating set $\{r\}$) and make some observations.  How are they similar and how are they different?  Can you reconcile the similarities and differences by thinking about the actions of each group?
\end{problem}

Hopefully, one of the things you noticed in the previous problem is that we can ``see" $R_4$ inside of $D_4$.  You may have used different colors in each case and maybe even labeled the vertices with different words, but the overall structure of $R_4$ is there nonetheless.

\begin{problem}\label{prob:R4_subgroup_D_4}
If you ignore the labels on the vertices and just pay attention to the configuration of arrows, it appears that there are two copies of the Cayley diagram for $R_4$ in the Cayley diagram for $D_4$.  Isolate these two copies by ignoring the edges that correspond to the generator $s$.  Now, paying close attention to the words that label the vertices from the original Cayley diagram for $D_4$, are either of these groups in their own right?
\end{problem}

Recall that the identity must be one of the elements included in a group.  If this didn't occur to you when doing the previous problem, you might want to go back and rethink your answer.  Just like in the previous problem, we can often ``see" smaller groups living inside larger groups.  These smaller groups are called \textbf{subgroups}.

\begin{definition}
Let $G$ be a group and let $H$ be a subset of $G$.  Then $H$ is a \textbf{subgroup} of $G$, written $H\leq G$, provided that $H$ is a group in its own right under the binary operation inherited from $G$.
\end{definition}

The phrase ``under the binary operation inherited from $G$" means that to combine two elements in $H$, we should treat the elements as if they were in $G$ and perform the binary operation of $G$.

In light of Problem~\ref{prob:R4_subgroup_D_4}, we would write $R_4\leq D_4$.  The second sub-diagram of the Cayley diagram for $D_4$ (using $\{r,s\}$ as the generating set) that resembles $R_4$ cannot be a subgroup because it does not contain the identity.  However, since it looks a lot like $R_4$, we call it a \textbf{clone} of $R_4$. For convenience, we also say that a subgroup is a clone of itself.

\begin{problem}\label{prob:informal_subgroup_criterion}
Let $G$ be a group and let $H\subseteq G$. If we wanted to determine whether $H$ is a subgroup of $G$, can we skip checking any of the axioms? Which axioms must we verify?
\end{problem}

Let's make the observations of the previous problem a bit more formal.

\begin{theorem}[Two Step Subgroup Test]\label{thm:subgroup_criterion}
Suppose $G$ is a group and $H$ is a nonempty subset of $G$.  Then $H\leq G$ if and only if (i) for all $h\in H$, $h^{-1} \in H$, as well, and (ii) $H$ is closed under the binary operation of $G$.
\end{theorem}

Notice that one of the hypotheses of Theorem~\ref{thm:subgroup_criterion} is that $H$ be nonempty.  This means that if we want to prove that a certain subset $H$ is a subgroup of a group $G$, then one of the things we must do is verify that $H$ is in fact nonempty. In light of this, the ``Two Step Subgroup Test" should probably be called the ``Three Step Subgroup Test".

As Theorems~\ref{thm:trivial_subgroup} and \ref{thm:improper_subgroup} will illustrate, there are a couple of subgroups that every group contains.

\begin{theorem}\label{thm:trivial_subgroup}
If $G$ is a group, then $\{e\}\leq G$.
\end{theorem}

The subgroup $\{e\}$ is referred to as the \textbf{trivial subgroup}.  All other subgroups are called \textbf{nontrivial}.

\begin{problem}
Let $G$ be a group. What does the Cayley diagram for the subgroup $\{e\}$ look like? What are you using as your generating set?
\end{problem}

Earlier, we referred to subgroups as being ``smaller."  However, our definition does not imply that this has to be the case.

\begin{theorem}\label{thm:improper_subgroup}
If $G$ is a group, then $G\leq G$.
\end{theorem}

We refer to subgroups that are not equal to the whole group as \textbf{proper subgroups}. If $H$ is a proper subgroup, then we may write $H<G$.

Recall Theorem~\ref{thm:subgroup_generated_by_S} that states that if $G$ is a group under $*$ and $S$ is a subset of $G$, then $\langle S\rangle$ is also a group under $*$.  Let's take this a step further.

\begin{theorem}\label{thm:smallest_subgroup_containing_S}
If $G$ is a group and $S\subseteq G$, then $\langle S\rangle \leq G$.  In particular, $\langle S\rangle$ is the smallest subgroup of $G$ containing $S$.
\end{theorem}

The subgroup $\langle S\rangle$ is called the \textbf{subgroup generated by $S$}.  In the special case when $S$ equals a single element, say $S=\{g\}$, then
\[
\langle g\rangle =\{g^k\mid k\in\mathbb{Z}\},
\]
which is called the (\textbf{cyclic}) \textbf{subgroup generated by $g$}. Every subgroup can be written in the ``generated by" form.  That is, if $H$ is a subgroup of a group $G$, then there always exists a subset $S$ of $G$ such that $\langle S\rangle=H$.  In particular, $\langle H\rangle=H$ for $H\leq G$, and as a special case, we have $\langle G\rangle=G$.

\begin{problem}
Consider $\Spin_{1\times 2}$ with generating set $\{s_{11}, s_{22},s_{12}\}$.  
\begin{enumerate}[label=\rm{(\alph*)}]
\item Find the Cayley diagram for the subgroup $\langle s_{11}\rangle$ inside the Cayley diagram for $\Spin_{1\times 2}$.  Identify all of the clones of $\langle s_{11}\rangle$ inside $\Spin_{1\times 2}$.
\item Find the Cayley diagram for the subgroup $\langle s_{11}, s_{22}\rangle$ inside the Cayley diagram of $\Spin_{1\times 2}$.  Identify the clones of $\langle s_{11}, s_{22}\rangle$ inside $\Spin_{1\times 2}$.
\end{enumerate}
\end{problem}

One of the benefits of Cayley diagrams is that they are useful for visualizing subgroups.  However, recall that if we change our set of generators, we might get a very different looking Cayley diagram.  The upshot of this is that we may be able to see a subgroup in one Cayley diagram for a given group, but not be able to see it in the Cayley diagram arising from a different generating set.

\begin{problem}
We currently have two different Cayley diagrams for $D_3$ (see Problems~\ref{prob:introducing_D3} and \ref{prob:alternate_D3}).  
\begin{enumerate}[label=\rm{(\alph*)}]
\item Can you find the Cayley diagram for the trivial subgroup $\langle e\rangle$ in either Cayley diagram for $D_3$?  Identify all of the clones of $\langle e\rangle$ in both Cayley diagrams for $D_3$.
\item Can you find the Cayley diagram for the subgroup $\langle r\rangle =R_3$ in either Cayley diagram for $D_3$?  If possible, identify all of the clones of $R_3$ in the Cayley diagrams for $D_3$.
\item Can you find the Cayley diagrams for $\langle s\rangle$ and $\langle s'\rangle$ in either Cayley diagram for $D_3$?  If possible, identify all of the clones of $\langle s\rangle$ and $\langle s'\rangle$ in the Cayley diagrams for $D_3$.
\end{enumerate}
\end{problem}

\begin{problem}\label{prob:subgroups_D4}
Consider $D_4$.  Let $h$ be the reflection of the square over the horizontal midline and let $v$ be the reflection over the vertical midline.  Which of the following are subgroups of $D_4$?  In each case, justify your answer.  If a subset is a subgroup, try to find a minimal generating set.  Also, determine whether you can see the subgroups in our Cayley diagram for $D_4$ with generating set $\{r,s\}$.
\begin{enumerate}[label=\rm{(\alph*)}]
\item $\{e, r^2\}$
\item $\{e,h\}$
\item $\{e, h, v\}$
\item\label{V4} $\{e, h, v, r^2\}$
\end{enumerate}
\end{problem}

Perhaps you recognized the set in part~\ref{V4} of the previous problem as being the Klein four-group $V_4$. It follows that $V_4\leq D_4$.














%Lots of groups have been given formal names (e.g., $D_4$, $R_4$, etc.).  However, not every group or subgroup has a name.  In this case, it's useful to have notation to refer to specific subgroups.

%\begin{definition}\label{def:subgroup_gen_by}
%Let $G$ be a group of actions and let $g_1,\ldots, g_n$ be distinct actions from $G$.  We define $\langle g_1,\ldots, g_n\rangle$ to be the smallest subgroup containing $g_1,\ldots, g_n$.  In this case, we call $\langle g_1,\ldots, g_n\rangle$ the \textbf{subgroup generated by} $g_1,\ldots, g_n$.
%\end{definition}
%
%For example, consider $r, s, s'\in D_3$ as defined in Exercises~\ref{prob:introducing_D3} and \ref{prob:alternate_D3}.  Then $D_3\langle r,s\rangle=\langle s, s'\rangle$.  Recall that $R_4$ is the subgroup of $D_4$ consisting of rotational symmetries of the square. In this case, $R_4=\langle r\rangle$.  Similarly, the group of rotations of an equilateral triangle is called $R_3$.  Then using the $r$ from $D_3$, we have $R_3=\langle r\rangle$, which is a subgroup of $D_3$.
%
%Note that in Definition~\ref{def:subgroup_gen_by}, we used a finite number of generators.  There's no reason we have to do this.  That is, we can consider groups/subgroups generated by infinitely many elements.

%\begin{problem}
%Suppose $\{g_1,\ldots,g_n\}$ is a generating set for a group $G$.
%\begin{enumerate}[label=\rm{(\alph*)}]
%\item Explain why $\{g^{-1}_1,\ldots,g^{-1}_n\}$ is also a generating set for $G$.
%\item How does the Cayley diagram for $G$ with generating set $\{g_1,\ldots,g_n\}$ compare to the Cayley diagram with generating set $\{g^{-1}_1,\ldots,g^{-1}_n\}$?
%\end{enumerate}
%\end{problem}

%\begin{problem}
%Consider $\Spin_{1\times 2}$.  
%\begin{enumerate}[label=\rm{(\alph*)}]
%\item Can you find the Cayley diagram for $\langle t_1\rangle$ inside the Cayley diagram for $\Spin_{1\times 2}$ that we previously constructed?
%\item Write down all the actions of the subgroup $\langle t_1, t_2\rangle$ by writing them as words in $t_1$ and $t_2$.  Can you find the Cayley diagram for $\langle t_1, t_2\rangle$ as a subgroup of $\Spin_{1\times 2}$?  Can you find a clone for $\langle t_1, t_2\rangle$ that is not the subgroup itself?
%\end{enumerate}
%\end{problem}
%
%One of the benefits of Cayley diagrams is that they are useful for visualizing subgroups.  However, recall that if we change our set of generators, we might get a very different looking Cayley diagram.  The upshot of this is that we may be able to see a subgroup in one Cayley diagram for a given group, but not be able to see it in the Cayley diagram arising from a different generating set.
%
%\begin{problem}
%We currently have two different Cayley diagrams for $D_3$ (see Exercises \ref{prob:introducing_D3} and \ref{prob:alternate_D3}).  
%\begin{enumerate}[label=\rm{(\alph*)}]
%\item Can you find the Cayley diagram for $\langle e\rangle$ as a subgroup of $D_3$?  Can you see it in both Cayley diagrams for $D_3$?  Can you find all the clones?
%\item Can you find the Cayley diagram for $\langle r\rangle =R_3$ as a subgroup of $D_3$?  Can you see it in both Cayley diagrams?  Can you find all the clones?
%\item Find the Cayley diagrams for $\langle s\rangle$ and $\langle s'\rangle$ as subgroups of $D_3$.  Can you see them in both Cayley diagrams for $D_3$?  Can you find all the clones?
%\end{enumerate}
%\end{problem}
%
%\begin{problem}\label{prob:subgroups_D4}
%Consider $D_4$.  Let $h$ be the action that reflects (i.e., flips over) the square over the horizontal midline and let $v$ be the action that reflects the square over the vertical midline.  Also, recall that $r^2$ is shorthand for the action $rr$ that does $r$ twice in a row.  Which of the following are subgroups of $D_4$?  In each case, justify your answer.  If a subset is a subgroup, try to find a minimal set of generators.  Also, determine whether you can see the subgroups in our Cayley diagram for $D_4$.
%\begin{enumerate}[label=\rm{(\alph*)}]
%\item $\{e, r^2\}$
%\item $\{e,h\}$
%\item $\{e, h, v\}$
%\item\label{V4} $\{e, h, v, r^2\}$
%\end{enumerate}
%\end{problem}
%
%The subgroup in Problem~\ref{prob:subgroups_D4}\ref{V4} is often referred to as the \textbf{Klein four-group} and is denoted by $V_4$.

%\begin{problem}\label{prob:V4}
%Draw the Cayley diagram for $V_4$ using $\{v,h\}$ as the generating set.
%\end{problem}

Let's introduce a group we haven't seen yet.  Define the \textbf{quaternion group} to be the group $Q_8=\{1,-1,i,-i,j,-j,k,-k\}$ having the Cayley diagram with generating set $\{i, j, -1\}$ given in Figure~\ref{fig:Q8}.  In this case, 1 is the identity of the group.

\tikzstyle{vert} = [circle, draw, fill=grey,inner sep=0pt, minimum size=6.5mm]
\tikzstyle{b} = [draw,very thick,blue,-stealth]
\tikzstyle{r} = [draw, very thick, red,-stealth]
\tikzstyle{g} = [draw, very thick, green, stealth-stealth]

\begin{figure}[!ht]
\centering
\begin{tikzpicture}[scale=1.5,auto]
\node (1) at (135:2) [vert] {\scriptsize $1$};
\node (i) at (45:2) [vert] {\scriptsize $i$};
\node (k) at (-45:2) [vert] {\scriptsize $k$};
\node (j) at (-135:2) [vert] {\scriptsize $j$};
\node (-1) at (135:1) [vert] {\scriptsize $-1$};
\node (-i) at (45:1) [vert] {\scriptsize $-i$};
\node (-k) at (-45:1) [vert] {\scriptsize $-k$};
\node (-j) at (-135:1) [vert] {\scriptsize $-j$};

\path[b] (1) to (i);
\path[b] (i) to (-1);
\path[b] (-1) to (-i);
\path[b] (-i) to (1);

\path[b] (-j) to (-k);
\path[b] (-k) to (j);
\path[b] (j) to (k);
\path[b] (k) to (-j);

\path[r] (-k) to (-i);
\path[r] (-i) to (k);
\path[r] (k) to (i);
\path[r] (i) to (-k);

\path[r] (1) to (j);
\path[r] (j) to (-1);
\path[r] (-1) to (-j);
\path[r] (-j) to (1);

\path[g] (1) to (-1);
\path[g] (j) to (-j);
\path[g] (i) to (-i);
\path[g] (k) to (-k);

\end{tikzpicture}
\caption{Cayley diagram for $Q_8$ with generating set $\{-1, i, j\}$.}\label{fig:Q8}
\end{figure}

Notice that I didn't mention what the actions actually do.  For now, let's not worry about that.  The relationship between the arrows and vertices tells us everything we need to know.  Also, let's take it for granted that $Q_8$ actually is a group.

\begin{problem}
Consider the Cayley diagram for $Q_8$ given in Figure~\ref{fig:Q8}.
\begin{enumerate}[label=\rm{(\alph*)}]
\item Which arrows correspond to which generators in our Cayley diagram for $Q_8$?
\item What is $i^2$ equal to?  That is, what element of $\{1,-1,i,-i,j,-j,k,-k\}$ is $i^2$ equal to?  How about $i^3$, $i^4$, and $i^5$?
\item What are $j^2$, $j^3$, $j^4$, and $j^5$ equal to?
\item What is $(-1)^2$ equal to?
\item What is $ij$ equal to?  How about $ji$?
\item Can you determine what $k^2$ and $ik$ are equal to?
\item Can you identify a generating set consisting of only two elements?  Can you find more than one?
\item What subgroups of $Q_8$ can you see in the Cayley diagram in Figure~\ref{fig:Q8}?
\item Find a subgroup of $Q_8$ that you cannot see in the Cayley diagram.
\end{enumerate}
\end{problem}

\begin{problem}
Consider $(\mathbb{R}^3,+)$, where $\mathbb{R}^3$ is the set of all 3-entry row vectors with real number entries (e.g., $(a,b,c)$ where $a,b,c\in\mathbb{R}$) and $+$ is ordinary vector addition.  It turns out that $(\mathbb{R}^3,+)$ is an abelian group with identity $(0,0,0)$.  
\begin{enumerate}[label=\rm{(\alph*)}]
\item Let $H$ be the subset of $\mathbb{R}^3$ consisting of vectors with first coordinate 0.  Is $H$ a subgroup of $\mathbb{R}^3$?  Prove your answer.
\item Let $K$ be the subset of $\mathbb{R}^3$ consisting of vectors whose entries sum to 0.  Is $K$ a subgroup of $\mathbb{R}^3$?  Prove your answer.
\item Construct a subset of $\mathbb{R}^3$ (different from $H$ and $K$) that is \emph{not} a subgroup of $\mathbb{R}^3$.
\end{enumerate}
\end{problem}

\begin{problem}\label{prob:nZ}
Consider the group $(\mathbb{Z},+)$ (under ordinary addition).
\begin{enumerate}[label=\rm{(\alph*)}]
\item Show that the even integers, written $2\mathbb{Z}:=\{2k\mid k\in\mathbb{Z}\}$, form a subgroup of $\mathbb{Z}$.
\item Show that the odd integers are not a subgroup of $\mathbb{Z}$.
\item Show that all subsets of the form $n\mathbb{Z}:=\{nk\mid k\in\mathbb{Z}\}$ for $n\in\mathbb{Z}$ are subgroups of $\mathbb{Z}$.
\item\label{prob:nZothers} Are there any other subgroups besides the ones listed in part (c)?  Explain your answer.
\item For $n\in \mathbb{Z}$, write the subgroup $n\mathbb{Z}$ in the ``generated by" notation.  That is, find a set $S$ such that $\langle S\rangle =n\mathbb{Z}$.  Can you find more than one way to do it?
\end{enumerate}
\end{problem}

\begin{problem}
Consider the group of symmetries of a regular octagon.  This group is denoted by $D_8$, where the operation is composition of actions.  The group $D_8$ consists of 16 elements (8 rotations and 8 reflections).  Let $H$ be the subset consisting of the following clockwise rotations: $0^\circ$, $90^\circ$, $180^\circ$, and $270^\circ$.  Determine whether $H$ is a subgroup of $D_8$ and justify your answer.
\end{problem}

\begin{problem}
Consider the groups $(\mathbb{R},+)$ and $(\mathbb{R}\setminus\{0\},\cdot)$.  Explain why $\mathbb{R}\setminus\{0\}$ is not a subgroup of $\mathbb{R}$ despite the fact that $\mathbb{R}\setminus\{0\}\subseteq\mathbb{R}$ and both are groups (under the respective binary operations).
\end{problem}

\begin{theorem}
If $G$ is an abelian group such that $H\leq G$, then $H$ is an abelian subgroup.
\end{theorem}

\begin{problem}
Is the converse of the previous theorem true?  If so, prove it.  Otherwise, provide a counterexample.
\end{problem}

As we've seen, some groups are abelian and some are not.  If $G$ is a group, then we define the \textbf{center} of $G$ to be
\[
Z(G):=\{z\in G\mid zg=gz\text{ for all } g\in G\}.
\]
Notice that if $G$ is abelian, then $Z(G)=G$.  However, if $G$ is not abelian, then $Z(G)$ will be a proper subset of $G$.  In some sense, the center of a group is a measure of how close $G$ is to being abelian.

\begin{theorem}
If $G$ is a group, then $Z(G)$ is an abelian subgroup of $G$.
\end{theorem}

\begin{problem}
Find the center of each of the following groups.
\begin{enumerate}[label=\rm{(\alph*)}]
\item $S_2$
\item $V_4$
\item $S_3$
\item $D_3$
\item $D_4$
\item $R_4$
\item $R_6$
\item $\Spin_{1\times 2}$
\item $Q_8$
\item $(\mathbb{Z},+)$
\item $(\mathbb{R}\setminus\{0\},\cdot)$
\end{enumerate}
\end{problem}

\end{section}

%%----------------------------%%

\begin{section}{Subgroup Lattices}

%%----------------------------%%

One of the goals of this section is to gain better understanding of the structure of groups by studying their subgroups. 

Suppose we wanted to find all of the subgroups of a finite group $G$. Theorems~\ref{thm:trivial_subgroup} and \ref{thm:improper_subgroup} tell us that $\{e\}$ and $G$ itself are subgroups of $G$, but there may be others.  Theorem~\ref{thm:subgroup_criterion} tells us that if we want to find other subgroups of $G$, we need to find nonempty subsets of $G$ that are closed and contain all the necessary inverses.  So, one method for finding subgroups would be to find all possible nonempty subsets of $G$ and then go about determining which subsets are subgroups by verifying whether a given subset is closed under inverses and closed under the operation of $G$. This is likely to be fairly time consuming.

Another approach would be to utilize the fact that every subgroup $H$ of $G$ has a generating set.  That is, if $H$ is a subgroup of a group $G$, then there always exists a subset $S$ of $G$ such that $\langle S\rangle=H$. Given a subset $S$ of $G$, $\langle S\rangle$ is guaranteed to be closed under inverses and the operation of the group $G$.  So, we could determine all of the subgroups of $G$ by generating groups with various subsets $S$ of $G$. Of course, one drawback is that it might take a bit of effort to determine what $\langle S\rangle$ actually is. Another drawback is that two different subsets $S$ and $T$ may generate the same subgroup.

Let's make this a bit more concrete by exploring an example.  Consider the group $R_4$.  What are the subgroups of $R_4$?  Since the order of $R_4$ is 4, we know that there are $2^4-1=15$ nonempty subsets of $R_4$.  Some of these are subgroups, but most of them are not.  Theorems~\ref{thm:trivial_subgroup} and \ref{thm:improper_subgroup} guarantee that $\{e\}$ and $R_4$ itself are subgroups of $R_4$.  That's 2 out of 15 so far.  Are there any others? Let's do an exhaustive search by playing with generating sets.  We can certainly be more efficient, but below we list all of the possible subgroups we can generate using subsets of $R_4$.  As you scan the list, you should take a moment to convince yourself that the list is accurate.
\begin{multicols}{2}
\begin{itemize}
\item[] $\langle e \rangle = \{e\}$
\item[] $\langle r \rangle  = \{e,r,r^2,r^3\}$
\item[] $\langle r^2 \rangle  = \{e,r^2\}$
\item[] $\langle r^3 \rangle  = \{e,r^3,r^2,r\}$
\item[] $\langle e,r \rangle  = \{e,r,r^2,r^3\}$
\item[] $\langle e,r^2 \rangle  = \{e,r^2\}$
\item[] $\langle e,r^3 \rangle  = \{e,r^3,r^2,r\}$
\item[] $\langle r,r^2 \rangle  = \{e,r,r^2,r^3\}$
\item[] $\langle r,r^3 \rangle  = \{e,r,r^2,r^3\}$
\item[] $\langle r^2,r^3 \rangle  = \{e,r,r^2,r^3\}$
\item[] $\langle e,r,r^2 \rangle  = \{e,r,r^2,r^3\}$
\item[] $\langle e,r,r^3 \rangle  = \{e,r,r^2,r^3\}$
\item[] $\langle e,r^2,r^3 \rangle  = \{e,r,r^2,r^3\}$
\item[] $\langle r,r^2,r^3 \rangle  = \{e,r,r^2,r^3\}$
\item[] $\langle e,r,r^2,r^3 \rangle = \{e,r,r^2,r^3\}$
\end{itemize}
\end{multicols}
\noindent Let's make a few observations.  Scanning the list, we see only three distinct subgroups:
\[
\{e\}, \{e,r^2\},\{e,r,r^2,r^3\}.
\]
Out of 15 nonempty subsets of $R_4$, only 3 subsets are subgroups. Our exhaustive search guarantees that these are the only subgroups of $R_4$.  It is also worth pointing out that if a subset contains either $r$ or $r^3$, then that subset generates all of $R_4$.  The reason for this is that $\{r\}$ and $\{r^3\}$ are each minimal generating sets for $R_4$. More generally, observe that if we increase the size of the generating subset using an element that was already contained in the subgroup generated by the set, then we don't get anything new.  For example, consider $\langle r^2\rangle=\{e,r^2\}$.  Since $e\in\langle r^2\rangle$, we don't get anything new by including $e$ in our generating set.  We can state this as a general fact.

\begin{theorem}
Let $G$ be a group and let $g_1,g_2,\ldots,g_n\in G$.  If $x\in\langle g_1,g_2,\ldots,g_n\rangle$, then $\langle g_1,g_2,\ldots,g_n\rangle = \langle g_1,g_2,\ldots,g_n,x\rangle$.
\end{theorem}

In the previous theorem, we are not claiming that $\{g_1,g_2,\ldots,g_n\}$ is a generating set for $G$---although this may be the case.  Instead, are simply making a statement about the subgroup $\langle g_1,g_2,\ldots,g_n\rangle$, whatever it may be.

We can capture the overall relationship between the subgroups of a group $G$ using a \textbf{subgroup lattice}.  Given a group $G$, the \textbf{lattice of subgroups} of $G$ is the partially ordered set whose elements are the subgroups of $G$ with the partial order relation being set inclusion. It is common to depict the subgroup lattice for a group using a \textbf{Hasse diagram}.  The Hasse diagram of subgroup lattice is drawn as follows:
\begin{enumerate}[label=\rm{(\arabic*)}]
\item Each subgroup $H$ of $G$ is a vertex.
\item Vertices corresponding to subgroups with smaller order are placed lower in the diagram than vertices corresponding to subgroups with larger order.  In particular, the vertex for $\{e\}$ is placed at the bottom of the diagram and the vertex for $G$ is placed at the top.
\item There is an edge going up from $H$ to $K$ if $H\leq K$ and there is no subgroup $L$ such that $H\leq L\leq K$ with $L\neq H,K$.
\end{enumerate}
Notice that there is an upward path of edges in the Hasse diagram from $H$ to $K$ iff $H\leq K$. For convenience we will not make a distinction between the subgroup lattice for a group $G$ and the corresponding Hasse diagram. 

The Hasse diagram for the subgroup lattice for $R_4$ is given in Figure~\ref{fig:latticeR4}.

\tikzstyle{b3} = [draw,very thick,black]

\begin{figure}[!ht]
\centering
\begin{tikzpicture}[scale=1.5,auto]
\node (a) at (0,0) {$\langle e\rangle=\{e\}$};
\node (b) at (0,2) {$\langle r^2\rangle=\{e,r^2\}$};
\node (c) at (0,4) {$\langle r\rangle=R_4$};

\path[b3] (a) to (b);
\path[b3] (b) to (c);
\end{tikzpicture}
\caption{Subgroup lattice for $R_4$.}
\label{fig:latticeR4}
\end{figure}

Let's see what we can do with $V_4=\{e,v,h,vh\}$.  Using an exhaustive search, we find that there are five subgroups:
\begin{itemize}
\item[] $\langle e \rangle = \{e\}$
\item[] $\langle h \rangle  = \{e,h\}$
\item[] $\langle v \rangle  = \{e,v\}$
\item[] $\langle vh \rangle  = \{e,vh\}$
\item[] $\langle v,h \rangle = \langle v,vh\rangle = \langle h, vh\rangle= \{e,v,h,vh\}=V_4$
\end{itemize}
For each subgroup above, we've used minimal generating sets to determine the subgroup.  The subgroup lattice for $V_4$ is given in Figure~\ref{fig:latticeV4}. Notice that there are no edges among $\langle v\rangle, \langle h\rangle$, and $\langle vh\rangle$.  The reason for this is that none of these groups are subgroups of each other.

\begin{figure}[!ht]
\centering
\begin{tikzpicture}[scale=1.5,auto]
\node (a) at (0,0) {$\langle e\rangle=\{e\}$};
\node (b) at (0,2) {$\langle h\rangle=\{e,h\}$};
\node (c) at (-2,2) {$\langle v\rangle=\{e,v\}$};
\node (d) at (2,2) {$\langle vh\rangle=\{e,vh\}$};
\node (e) at (0,4) {$\langle v,h\rangle=V_4$};

\path[b3] (a) to (b);
\path[b3] (a) to (c);
\path[b3] (a) to (d);
\path[b3] (b) to (e);
\path[b3] (c) to (e);
\path[b3] (d) to (e);
\end{tikzpicture}
\caption{Subgroup lattice for $V_4$.}
\label{fig:latticeV4}
\end{figure}

The next two theorems provide some further insight into the overall structure of subgroups of a group.

\begin{theorem}\label{thm:subgroup_intersection}
If $G$ is a group such that $H,K\leq G$, then $H\cap K\leq G$. Moreover, $H\cap K$ is the largest subgroup contained in both $H$ and $K$.
\end{theorem}

It turns out that we cannot simply replace ``intersection" with ``union" in the previous theorem

\begin{problem}
Provide an example of a group $G$ and subgroups $H$ and $K$ such that $H\cup K$ is not a subgroup of $G$.
\end{problem}

\begin{theorem}\label{thm:subgroup_generated_union}
If $G$ is a group such that $H,K\leq G$, then $\langle H\cup K\rangle\leq G$. Moreover, $\langle H\cup K\rangle\leq G$ is the smallest subgroup containing both $H$ and $K$.
\end{theorem}

Theorems~\ref{thm:subgroup_intersection} and \ref{thm:subgroup_generated_union} justify the use of the word ``lattice" in ``subgroup lattice".  In general, a partially ordered set in which every two elements have a unique \textbf{meet} (also called a \textbf{greatest lower bound} or \textbf{infimum}) and a unique \textbf{join} (also called a \textbf{least upper bound} or \textbf{supremum}).  In the case of a subgroup lattice for a group $G$, the meet of subgroups $H$ and $K$ is $H\cap K$ and the join is $\langle H\cup K\rangle$.  Figure~\ref{fig:diamond} illustrates the meet (Theorem~\ref{thm:subgroup_intersection}) and join (Theorem~\ref{thm:subgroup_generated_union}) in the case when $H$ and $K$ are not comparable.

\begin{figure}[!ht]
\centering
\begin{tikzpicture}[scale=1.5,auto]
\node (a) at (0,0) {$H\cap K$};
\node (b) at (-2,2) {$H$};
\node (c) at (2,2) {$K$};
\node (d) at (0,4) {$\langle H\cup K\rangle$};
\path[b3] (a) to (-.75,.75);
\path[b3,dashed] (-.75,.75) to (-1.25,1.25);
\path[b3] (-1.25,1.25) to (b);
\path[b3] (a) to (.75,.75);
\path[b3,dashed] (.75,.75) to (1.25,1.25);
\path[b3] (1.25,1.25) to (c);
\path[b3] (c) to (1.25,2.75);
\path[b3,dashed] (1.25,2.75) to (.75,3.25);
\path[b3] (.75,3.25) to (d);
\path[b3] (b) to (-1.25,2.75);
\path[b3,dashed] (-1.25,2.75) to (-.75,3.25);
\path[b3] (-.75,3.25) to (d);
\end{tikzpicture}
\caption{Meet and join for subgroups $H$ and $K$.}
\label{fig:diamond}
\end{figure}

In the next few problems, you are asked to create subgroup lattices.  As you do this, try to minimize the amount of work it takes to come up with all the subgroups.

\begin{problem}
Find all the subgroups of $R_5=\{e,r,r^2,r^3,r^4\}$ (where $r$ is clockwise rotation of a regular pentagon by $72^{\circ}$) and then draw the subgroup lattice for $R_5$.
\end{problem}

\begin{problem}
Find all the subgroups of $R_6=\{e,r,r^2,r^3,r^4,r^5\}$ (where $r$ is clockwise rotation of a regular hexagon by $60^{\circ}$) and then draw the subgroup lattice for $R_6$.
\end{problem}

\begin{problem}\label{prob:latticeD3}
Find all the subgroups of $D_3=\{e,r,r^2,s,sr,sr^2\}$ (where $r$ and $s$ are the usual symmetries of an equilateral triangle) and then draw the subgroup lattice for $D_3$.
\end{problem}

\begin{problem}
Find all the subgroups of $S_3=\langle s_1, s_2\rangle$ (where $s_1$ is the action is that swaps the positions of the first and second coins and $s_2$ is the action that swaps the second and third coins; see Problem~\ref{prob:revisiting_S3}) and then draw the subgroup lattice for $S_3$. How does your lattice compare to the one in Problem~\ref{prob:latticeD3}? You should look back at parts~\ref{cayley:altD3} and \ref{cayley:S3} of Problem~\ref{prob:make_Cayley_diagrams} and ponder what just happened.
\end{problem}

\begin{problem}
Find all the subgroups of $D_4=\{e,r,r^2,r^3,s,sr,sr^2,sr^3\}$ (where $r$ and $s$ are the usual symmetries of a square) and then draw the subgroup lattice for $D_4$.
\end{problem}

The last problem in this section is a step up in difficulty.

\begin{problem}
Find all the subgroups of $Q_8=\{1,-1,i,-i,j,-j,k,-k\}$ and then draw the subgroup lattice for $Q_8$.
\end{problem}

\end{section}

%%----------------------------%%

\begin{section}{Isomorphisms}

%%----------------------------%%

As we have been exploring various groups, I'm sure you've noticed that some groups seem to look and behave the same.  For example, if we choose the same colors for our arrows and ignore the labels on the vertices, the Cayley diagram for $D_3$ with generating set $\{s,s'\}$ looks the same as the Cayley diagram for $S_3$ with generating set $\{s_1, s_2\}$. That is, if we pick the appropriate colors and set the Cayley diagram for $D_3$ (with generating set $\{s,s'\}$) on top of the Cayley diagram for $S_3$ (with generating set $\{s_1, s_2\}$) such that the identities match up, then the two Cayley diagrams are identical up to relabeling the rest of the vertices.  Figure~\ref{fig:D3andS3} should make this clear.  This act of matching up the Cayley diagrams establishes a correspondence between the elements of the two groups:
\begin{align*}
e & \mapsto e\\
s & \mapsto s_1\\
s' &\mapsto s_2\\
ss' &\mapsto s_1s_2\\
s's & \mapsto s_2s_1\\
ss's & \mapsto s_1s_2s_1
\end{align*}
Notice that each correspondence is compatible with the correspondence of the generators, namely: $s \mapsto s_1$ and $s' \mapsto s_2$. Given this correspondence, it should not be surprising that the subgroup lattices for $D_3$ and $S_3$ have the same structure.


\tikzstyle{vert} = [circle, draw, fill=grey,inner sep=0pt, minimum size=8.5mm]
\tikzstyle{b2} = [draw,very thick,blue,stealth-stealth]
\tikzstyle{r2} = [draw, very thick, red,stealth-stealth]

\begin{figure}[!ht]
\centering
\begin{tikzpicture}[scale=1.75,auto]
\node (s's) at (-30:1.5) [vert] {{\scriptsize $s's$}};
\node (s) at (30:1.5) [vert] {\scriptsize {$s$}};
\node (e) at (90:1.5) [vert] {{\scriptsize $e$}};
\node (s') at (150:1.5) [vert] {{\scriptsize $s'$}};
\node (ss') at (210:1.5) [vert] {{\scriptsize $ss'$}};
\node (ss's) at (-90:1.5) [vert] {{\scriptsize $ss's$}};
\path[r2] (e) to (s);
\path[b2] (e) to (s');
\path[r2] (s') to (ss');
\path[b2] (ss') to (ss's);
\path[b2] (s) to (s's);
\path[r2] (s's) to (ss's);
\node (D3) at (0,0) {$D_3$};
\end{tikzpicture}
\hspace{1cm}
\begin{tikzpicture}[scale=1.75,auto]
\node (s's) at (-30:1.5) [vert] {{\scriptsize $s_2s_1$}};
\node (s) at (30:1.5) [vert] {\scriptsize {$s_1$}};
\node (e) at (90:1.5) [vert] {{\scriptsize $e$}};
\node (s') at (150:1.5) [vert] {{\scriptsize $s_2$}};
\node (ss') at (210:1.5) [vert] {{\scriptsize $s_1s_2$}};
\node (ss's) at (-90:1.5) [vert] {{\scriptsize $s_1s_2s_1$}};
\path[r2] (e) to (s);
\path[b2] (e) to (s');
\path[r2] (s') to (ss');
\path[b2] (ss') to (ss's);
\path[b2] (s) to (s's);
\path[r2] (s's) to (ss's);
\node (S3) at (0,0) {$S_3$};
\end{tikzpicture}
\caption{Cayley diagrams for $D_3$ and $S_3$ with generating sets $\{s,s'\}$ and $\{s_1,s_2\}$, respectively.}
\label{fig:D3andS3}
\end{figure}

Perhaps you've also noticed that the Cayley diagram for $R_4$ with generating set $\{r\}$ looks like the Cayley diagram for the subgroup $\langle j\rangle=\{\pm 1,\pm j\}$ with generating set $\{j\}$ in $Q_8$.  We've depicted the correspondence in Figure~\ref{fig:<j>isoR4}. Similarly, the Cayley diagram for $S_2$ with generating set $\{s\}$ looks like the Cayley diagram for the subgroup $\langle -1\rangle=\{\pm 1\}$ with generating set $\{-1\}$ in $Q_8$ (see Figure~\ref{fig:isoS2}), and both of these Cayley diagrams look like the Cayley diagram for the subgroup $\langle v\rangle =\{e,v\}$ with generating set $\{v\}$ in $V_4$.  The goal of this section is to formalize this phenomenon by introducing the notion of an \textbf{isomorphism}. First, let's develop a little more intuition.

\tikzstyle{correspondence} = [draw, very thick, black, dashed, stealth-stealth]
\tikzstyle{vert} = [circle, draw, fill=grey,inner sep=0pt, minimum size=6.5mm]
\tikzstyle{b} = [draw,very thick,blue,-stealth]
\tikzstyle{r} = [draw, very thick, red,-stealth]
\tikzstyle{g} = [draw, very thick, green,stealth-stealth]

\begin{figure}[!ht]
\centering
\begin{tikzpicture}[scale=1.5,auto]
\node (1) at (135:2.2) [vert] {\scriptsize $1$};
\node (i) at (45:2.2) [vert] {\scriptsize $i$};
\node (k) at (-45:2.2) [vert] {\scriptsize $k$};
\node (j) at (-135:2.2) [vert] {\scriptsize $j$};
\node (-1) at (135:1) [vert] {\scriptsize $-1$};
\node (-i) at (45:1) [vert] {\scriptsize $-i$};
\node (-k) at (-45:1) [vert] {\scriptsize $-k$};
\node (-j) at (-135:1) [vert] {\scriptsize $-j$};

\path[b] (1) to (i);
\path[b] (i) to (-1);
\path[b] (-1) to (-i);
\path[b] (-i) to (1);

\path[b] (-j) to (-k);
\path[b] (-k) to (j);
\path[b] (j) to (k);
\path[b] (k) to (-j);

\path[r] (-k) to (-i);
\path[r] (-i) to (k);
\path[r] (k) to (i);
\path[r] (i) to (-k);

\path[r] (1) to (j);
\path[r] (j) to (-1);
\path[r] (-1) to (-j);
\path[r] (-j) to (1);

\path[g] (1) to (-1);
\path[g] (j) to (-j);
\path[g] (i) to (-i);
\path[g] (k) to (-k);
\node (Q8) at (0,0) {$Q_8$};

\node (r) at (-3,.75) [vert] {\scriptsize $r$};
\node (e) at (-4.5,.75) [vert] {\scriptsize $e$};
\node (r3) at (-4.5,-.75) [vert] {\scriptsize $r^3$};
\node (r2) at (-3,-.75) [vert] {\scriptsize $r^2$};
\path[r] (e) to (r);
\path[r] (r) to (r2); 
\path[r] (r2) to (r3);
\path[r]  (r3) to (e);
\node (R4) at (-3.75,0) {$R_4$};

\path[correspondence] (1) to [bend right=60] (e);
\path[correspondence] (j) to [bend right=30] (r);
\path[correspondence] (-1) to [bend right=30] (r2);
\path[correspondence] (-j) to [bend left=30] (r3);

\end{tikzpicture}
\caption{Correspondence between $R_4=\langle r\rangle$ and $\langle j\rangle\leq Q_8$.}\label{fig:<j>isoR4}
\end{figure}


\begin{figure}[!ht]
\centering
\begin{tikzpicture}[scale=1.5,auto]
\node (1) at (135:2.2) [vert] {\scriptsize $1$};
\node (i) at (45:2.2) [vert] {\scriptsize $i$};
\node (k) at (-45:2.2) [vert] {\scriptsize $k$};
\node (j) at (-135:2.2) [vert] {\scriptsize $j$};
\node (-1) at (135:1) [vert] {\scriptsize $-1$};
\node (-i) at (45:1) [vert] {\scriptsize $-i$};
\node (-k) at (-45:1) [vert] {\scriptsize $-k$};
\node (-j) at (-135:1) [vert] {\scriptsize $-j$};

\path[b] (1) to (i);
\path[b] (i) to (-1);
\path[b] (-1) to (-i);
\path[b] (-i) to (1);

\path[b] (-j) to (-k);
\path[b] (-k) to (j);
\path[b] (j) to (k);
\path[b] (k) to (-j);

\path[r] (-k) to (-i);
\path[r] (-i) to (k);
\path[r] (k) to (i);
\path[r] (i) to (-k);

\path[r] (1) to (j);
\path[r] (j) to (-1);
\path[r] (-1) to (-j);
\path[r] (-j) to (1);

\path[g] (1) to (-1);
\path[g] (j) to (-j);
\path[g] (i) to (-i);
\path[g] (k) to (-k);

\node (e) at (-3,0) [vert] {\scriptsize $e$};
\node (s) at (-5,0) [vert] {\scriptsize $s$};
\path[g] (e) to (s);

\node (Q8) at (0,0) {$Q_8$};
\node (S2) at (-4,.3) {$S_2$};

\path[correspondence] (1) to [bend left=30] (e);
\path[correspondence] (-1) to [bend left=50] (s);

\end{tikzpicture}
\caption{Correspondence between $S_2=\langle s\rangle$ and $\langle -1\rangle\leq Q_8$.}\label{fig:isoS2}
\end{figure}


\begin{problem}
We have seen two different Cayley diagrams for $D_3$, one with generating set $\{s,r\}$ and one with generating set $\{s,s'\}$. As Figure~\ref{fig:D3andS3} illustrates, the Cayley diagrams for $D_3$ with generating set $\{s,s'\}$ and $S_3$ with generating set $\{s_1,s_2\}$ look the same.  Find a different generating set for $S_3$ such that the corresponding Cayley diagram looks like the Cayley diagram for $D_3$ with generating set $\{s,r\}$.
\end{problem}

\begin{problem}

\end{problem}

$s$ and $-1$ would correspond to each other.  In other words, the two Cayley diagrams are identical up to relabeling the vertices.

Suppose $(G_1,*)$ and $(G_2,\odot)$ are two groups.  Recall that $G_1$ and $G_2$ are isomorphic, written $G_1\cong G_2$, provided that we can choose generating sets for $G_1$ and $G_2$, respectively, so that the Cayley diagrams for both groups are identical (ignoring the labels on the vertices).  When two groups are isomorphic, it means that they have identical structure up to relabeling the names of the elements of the group.

One consequence of two groups being isomorphic is that there is a one-to-one correspondence between the elements of the group.  This correspondence is referred to as an isomorphism.  In other words, an isomorphism is a one-to-one and onto function that preserves the structure of the two groups.  

Having an isomorphism between two groups immediately implies that they have the same order, i.e., $|G_1|=|G_2|$ (see Theorem~\ref{thm:iso_same_order}).  However, it is extremely important to remember that two groups having the same order does \emph{not} imply that the two groups are isomorphic.  Said another way, having a one-to-one correspondence between two groups does not imply that the two groups are isomorphic.  They must also have the same structure!

\begin{exercise}
Provide an example of two groups that have the same order but are not isomorphic.
\end{exercise}

After we introduced groups tables, we also discussed the fact that $G_1\cong G_2$ exactly when we can arrange the rows and columns and color elements in such a way that the colorings for the two group tables agree (see Problem~\ref{prob:iso_same_group_table}).  The upshot of this is that if $G_1\cong G_2$, then
\begin{quotation}
\emph{the product of corresponding elements yields the corresponding result.}
\end{quotation}
This is the essence of what it means for two groups to have the same structure.  

Let's try to make a little more sense of this.  Suppose that $G_1\cong G_2$ and imagine we have arranged the rows and columns of their respective group tables and colored the elements in such a way that the colorings for the two group tables agree.  Now, let $x,y\in G_1$.  Then these two elements have corresponding elements in the group table for $G_2$, say $x'$ and $y'$, respectively.  In other words, $x$ and $x'$ have the same color while $y$ and $y'$ have the same color.  Since $G_1$ is closed under its binary operation $*$, there exists $z\in G_1$ such that $z=x*y$.  There must exist a $z'\in G_2$ such that $z'$ has the same color as $z$.  What must be true of $x'\odot y'$?  Since the two tables exhibit the same color pattern, it must be the case that $z'=x'\odot y'$.  This is what it means for the product of corresponding elements to yield the corresponding result.  Figure~\ref{fig:isoGroupTables} depicts this phenomenon for group tables.

\begin{figure}
\begin{center}
\begin{tabu}{c|[2pt]ccc|c|c}
$*$                & & & & \cellcolor{green}$y$  & \\ \tabucline[2pt]{-}
                   & & & &                       & \\ \hline
\cellcolor{red}$x$ & & & & \cellcolor{orange}$z$ & \\ \hline
                   & & & &                      & \\
                   & & & &                      &
\end{tabu}
\hspace{1cm}
$\longleftrightarrow$
\hspace{1cm}
\begin{tabu}{c|[2pt]ccc|c|c}
$\odot$                & & & & \cellcolor{green}$y'$  & \\ \tabucline[2pt]{-}
                   & & & &                       & \\ \hline
\cellcolor{red}$x'$ & & & & \cellcolor{orange}$z'$ & \\ \hline
                   & & & &                      & \\
                   & & & &                      &
\end{tabu}
\end{center}
\caption{}\label{fig:isoGroupTables}
\end{figure}

We can describe the isomorphism between $G_1$ and $G_2$ using a function.  Let $\phi:G_1\to G_2$ be the one-to-one and onto function that maps elements of $G_1$ to their corresponding elements in $G_2$.  Then $\phi(x)=x'$, $\phi(y)=y'$, and $\phi(z)=z'$.  Since $z'=x'\odot y'$, we can obtain
\[
\phi(x*y)=\phi(z)=z'=x'\odot y'=\phi(x)\odot \phi(y).
\]
In summary, it must be the case that 
\[
\phi(x*y)=\phi(x)\odot \phi(y).
\]
We are now prepared to state a formal definition of what it means for two groups to be isomorphic.

\begin{definition}\label{def:iso}
Let $(G_1,*)$ and $(G_2,\odot)$ be two groups.  Then $G_1$ is \textbf{isomorphic} to $G_2$, written $G_1\cong G_2$, if and only if there exists a one-to-one and onto function $\phi:G_1\to G_2$ such that
\begin{equation}\label{hom_property}
\phi(x*y)=\phi(x)\odot \phi(y).
\end{equation}
The function $\phi$ is referred to as an \textbf{isomorphism}.  Equation \ref{hom_property} is often referred to as the \textbf{homomorphic property}.
\end{definition}

You should definitely take a few minutes to convince yourself that the above definition agrees with our previous informal approach to isomorphisms.  For those of you that have had linear algebra, notice that our homomorphic property looks a lot like the requirement for a function on vector spaces to be a linear transformation.  Linear transformations preserve the algebraic structure of vector spaces while the homomorphic property is preserving the algebraic structure of groups.

We've seen several instances of two groups being isomorphic, but now that we have a formal definition, we can open the door to more possibilities.

\begin{problem}
Consider the groups $(\mathbb{R},+)$ and $(\mathbb{R}^+,\cdot)$, where $\mathbb{R}^+$ is the set of positive real numbers.  It turns out that these two groups are isomorphic, but this would be difficult to discover using our previous techniques because the groups are infinite.  Define $\phi:\mathbb{R}\to \mathbb{R}^+$ via $\phi(r)=e^r$ (where $e$ is the natural base, not the identity). Prove that $\phi$ is an isomorphism.
\end{problem}

\begin{exercise}
For each of the following pairs of groups, determine whether the given function is an isomorphism from the first group to the second group.
\begin{enumerate}[label=\rm{(\alph*)}]
\item $(\mathbb{Z},+)$ and $(\mathbb{Z},+)$, $\phi(n)=n+1$.
\item $(\mathbb{Z},+)$ and $(\mathbb{Z},+)$, $\phi(n)=-n$.
\item $(\mathbb{Q},+)$ and $(\mathbb{Q},+)$, $\phi(x)=x/2$.
\end{enumerate}

\end{exercise}

\begin{problem}
Show that the groups $(\mathbb{Z},+)$ and $(2\mathbb{Z},+)$ are isomorphic.
\end{problem}

Perhaps one surprising consequence of the previous problem is that when dealing with infinite groups, a group can have a proper subgroup that it is isomorphic to.  Of course, this never happens with finite groups.

Once we know that two groups are isomorphic, there are lots of interesting things we can say.  The next theorem tells us that isomorphisms map the identity element of one group to the identity of the second group.  It was already clear that this was the case using our informal definition of isomorphic.  Prove the next theorem using Definition~\ref{def:iso}

\begin{theorem}\label{thm:hom_id}
Suppose $\phi:G_1\to G_2$ is an isomorphism from the group $(G_1,*)$ to the group $(G_2,\odot)$.  If $e$ and $e'$ are the identity elements of $G_1$ and $G_2$, respectively, then $\phi(e)=e'$.
\end{theorem}

\begin{theorem}\label{thm:hom_inverse}
Suppose $\phi:G_1\to G_2$ is an isomorphism from the group $(G_1,*)$ to the group $(G_2,\odot)$.   Then $\phi(g^{-1})=[\phi(g)]^{-1}$.
\end{theorem}

\begin{theorem}
Suppose $\phi:G_1\to G_2$ is an isomorphism from the group $(G_1,*)$ to the group $(G_2,\odot)$. If $G_1$ is abelian, then $G_2$ is abelian.
\end{theorem}

\begin{theorem}
Suppose $\phi:G_1\to G_2$ is an isomorphism from the group $(G_1,*)$ to the group $(G_2,\odot)$. Then the function $\phi^{-1}:G_2\to G_1$ is an isomorphism.
\end{theorem}

\begin{theorem}
Suppose $\phi:G_1\to G_2$ and $\psi:G_2\to G_3$ are isomorphisms from the groups $(G_1,*)$ to $(G_2,\odot)$ and $(G_2,\odot)$ to $(G_3,\star)$, respectively. Then the composite function $\psi\circ\phi$ is an isomorphism of $G_1$ and $G_3$.
\end{theorem}

\begin{theorem}
Let $\mathcal{G}$ be any nonempty collection of groups.  Then the relation $\cong$ of being isomorphic is an equivalence relation.
\end{theorem}

\begin{theorem}
Suppose $\phi:G_1\to G_2$ is an isomorphism from the group $(G_1,*)$ to the group $(G_2,\odot)$.  If $H\leq G_1$, then $\phi(H)\leq G_2$, where
\[
\phi(H):=\{y\in G_2\mid \text{there exists } h\in H\text{ such that }\phi(h)=y\}. 
\]
Note that $\phi(H)$ is called the \textbf{image} of $H$.
\end{theorem}

\begin{theorem}
Suppose $(G,*)$ is a group and let $g\in G$.  Define $\phi_g:G\to G$ via $\phi(x)=gxg^{-1}$.  Then $\phi_g$ is an isomorphism from $G$ to $G$.  Note that the map $\phi_g$ is called \textbf{conjugation} by $g$.
\end{theorem}

Now that you've proved the above theorems, it's a good idea to review the key themes.  If you were really paying attention, you may have noticed that in a few of the proofs, we did not use the fact that the function was one-to-one and onto despite assuming that the function was an isomorphism.

\begin{problem}
For which of the recent theorems could we remove the assumption that the function is one-to-one and onto and only assume that it satisfies the homomorphic property?  Such functions are called \textbf{homomorphisms} and will be the subject of a future chapter.
\end{problem}


%By now you've probably seen enough examples of Cayley diagrams to witness some patterns appearing over and over again.    One of the things you've probably noticed is that parts of some Cayley diagrams look just like parts of other Cayley diagrams.
%
%Recall from Exercise~\ref{exer:introducing_S2} that $S_2$ is the group that acts on two coins by swapping their positions (but not flipping them over).  We defined $s$ to be the action that swaps the left and right coins and as usual $e$ is the do-nothing action.  The Cayley diagram for $S_2$ with generator $s$ is given in Figure~\ref{fig:S2}. 
%
%\tikzstyle{vert} = [circle, draw, fill=grey,inner sep=0pt, minimum size=6.5mm]
%\tikzstyle{twoway} = [draw, very thick, blue, stealth-stealth]
%
%\begin{figure}
%\centering
%\begin{tikzpicture}[scale=1.5,auto]
%\node (e) at (0,0) [vert] {\scriptsize $e$};
%\node (s) at (2,0) [vert] {\scriptsize $s$};
%\path[twoway] (e) to (s);
%\end{tikzpicture}
%\caption{Cayley diagram for $S_2$ with generator $s$.}\label{fig:S2}
%\end{figure}
%
%If you look back at all the Cayley diagrams you've encountered, you'll notice that many of them contained chunks that resemble the Cayley diagram for $S_2$ with generator $s$.  In particular, in the Cayley diagrams for $\Spin_{1\times 2}$, $D_3$, $D_4$, and $Q_8$ that we've seen, it is easy to identify the portions that ``look like" $S_2$.  For example, if you isolate the Cayley diagram for the subgroup $\langle -1\rangle=\{1,-1\}$ in $Q_8$, we see that it looks just like the Cayley diagram for $S_2$, except the labels are not identical.  The clones of the subgroup $\langle -1\rangle=\{1,-1\}$ in $Q_8$ look like $S_2$, as well, but they do not contain the do-nothing action.  
%
%The one thing that is different about the Cayley diagram for $S_2$ and the Cayley diagram for $\langle -1\rangle$ is that the labels are different.  If we set the Cayley diagram for $S_2$ on top of the Cayley diagram for $\langle -1\rangle$ such that the do-nothing actions match up, then $s$ and $-1$ would correspond to each other.  In other words, the two Cayley diagrams are identical up to relabeling the vertices.
%
%In this case, we say that $S_2$ and the subgroup $\langle -1\rangle$ of $Q_8$ are \textbf{isomorphic} under the correspondence $e\leftrightarrow 1$ and $s\leftrightarrow -1$.  This one-to-one correspondence between the two groups is called an \textbf{isomorphism}, which is depicted in Figure~\ref{fig:isoS2}. Note that I've recolored the arrow in $S_2$ so that it matches the corresponding arrow color of $\langle -1\rangle$.  This isn't necessary, but it makes the correspondence more obvious.
%
%\tikzstyle{correspondence} = [draw, very thick, black, dashed, stealth-stealth]
%\tikzstyle{vert} = [circle, draw, fill=grey,inner sep=0pt, minimum size=6.5mm]
%
%\begin{figure}[!ht]
%\centering
%\begin{tikzpicture}[scale=1.5,auto]
%\node (1) at (135:2) [vert] {\scriptsize $1$};
%\node (i) at (45:2) [vert] {\scriptsize $i$};
%\node (k) at (-45:2) [vert] {\scriptsize $k$};
%\node (j) at (-135:2) [vert] {\scriptsize $j$};
%\node (-1) at (135:1) [vert] {\scriptsize $-1$};
%\node (-i) at (45:1) [vert] {\scriptsize $-i$};
%\node (-k) at (-45:1) [vert] {\scriptsize $-k$};
%\node (-j) at (-135:1) [vert] {\scriptsize $-j$};
%
%\path[b] (1) to (i);
%\path[b] (i) to (-1);
%\path[b] (-1) to (-i);
%\path[b] (-i) to (1);
%
%\path[b] (-j) to (-k);
%\path[b] (-k) to (j);
%\path[b] (j) to (k);
%\path[b] (k) to (-j);
%
%\path[r] (-k) to (-i);
%\path[r] (-i) to (k);
%\path[r] (k) to (i);
%\path[r] (i) to (-k);
%
%\path[r] (1) to (j);
%\path[r] (j) to (-1);
%\path[r] (-1) to (-j);
%\path[r] (-j) to (1);
%
%\path[g] (1) to (-1);
%\path[g] (j) to (-j);
%\path[g] (i) to (-i);
%\path[g] (k) to (-k);
%
%\node (e) at (3,0) [vert] {\scriptsize $e$};
%\node (s) at (5,0) [vert] {\scriptsize $s$};
%\path[g] (e) to (s);
%
%\node (Q8) at (0,0) {$Q_8$};
%\node (S2) at (4,-.5) {$S_2$};
%
%\path[correspondence] (1) to [bend left=60] (e);
%\path[correspondence] (-1) to [bend left=60] (s);
%
%\end{tikzpicture}
%\caption{Isomorphism between $\langle -1\rangle\leq Q_8$ and $S_2$.}\label{fig:isoS2}
%\end{figure}
%
%What this means is that these two groups have the same structure and characteristics.  Or, in other words, these two groups essentially do the ``same kind" of thing.  Clearly, the two do-nothing actions behave the same way.  Also, $s$ and $-1$ both have the property that doing the action twice results in having done nothing (i.e., each element is its own reverse).  Since there are only two elements, there isn't anything else to check.  In groups with more elements, things can get much more complicated.  
%
%It is important to point out that $S_2$ and $\langle -1\rangle$ (in $Q_8$) are not equal.  But they have the same structure.  Identifying when two groups have the same structure (i.e., isomorphic) is an important pursuit in group theory.
%
%If you look at the original Cayley diagram for $\Spin_{1\times 2}$ (with generators $s, t_1, t_2$), we can see three subgroups that look like $S_2$; namely $\langle s\rangle$, $\langle t_1\rangle$, and $\langle t_2\rangle$.  Each of these three subgroups is isomorphic to $S_2$.  
%
%There is one serious potential for confusion here.  Notice that there is an $s$ in $S_2$ and an $s$ in $\Spin_{1\times 2}$.  Despite having identical names, they are not the same element.  Since we only have 26 letters in our alphabet this sort of thing is unavoidable.  Under the isomorphism between $S_2$ and the subgroup $\langle s\rangle$ in $\Spin_{1\times 2}$, the two elements with the same name match up.  That is, these two elements are the ones in each group with the same behavior.
%
%\begin{exercise}
%Can you find any other subgroups or groups that are isomorphic to $S_2$?
%\end{exercise}
%
%Let's write down an official definition of isomorphic.
%
%\begin{definition}
%Let $G$ and $G'$ be two groups.  We say that $G$ and $G'$ are \textbf{isomorphic} if there exist generating sets $S$ and $S'$ for $G$ and $G'$, respectively, such that the corresponding Cayley diagrams are identical where we ignore the labels on the vertices and recolor the edges if necessary.  In this case, we write $G\cong G'$.  Otherwise, we say that $G$ and $G'$ are not isomorphic.  If $G$ and $G'$ are isomorphic, then the one-to-one correspondence determined by matching up the corresponding generators and respecting arrow paths is called an \textbf{isomorphism}.
%\end{definition}
%
%The last sentence in the definition above might be a bit much to handle at the moment, but as we construct more examples, the concept should become clear.  The general idea is to take two identical Cayley diagrams (ignoring labels) for $G$ and $G'$ and then set one on top of the other so that the vertices and arrows of the same color match up.  This should be done so that the do-nothing actions correspond to each other.  Then it becomes clear which actions in $G$ correspond to which actions in $G'$.  There might be many ways to do this.
%
%Consider the group $R_4$ with generator $r$ (rotation by $90^\circ$ clockwise).  Now, take a look at the Cayley diagram for $Q_8$ with generators $i, j, -1$.  It should be easy to convince yourself that $R_4$ is isomorphic to both $\langle i\rangle=\{1,i,-i,-1\}$ and $\langle j\rangle=\{1,j,-j,-1\}$.  However, you have to do some rearranging of one of the diagrams to set one on top of the other.  Let's just focus on $\langle i\rangle$.  
%
%How do $R_4$ and $\langle i\rangle$ match up?  We want to pair elements in each group with an element in the other group that has the same behavior.  Clearly, $e$ and $1$ match up since these are the two do-nothing actions.  Also, the reason why we noticed these two groups were isomorphic is because their Cayley diagrams looked the same.  Since each Cayley diagram only had one arrow type determined by $r$ and $i$, we should pair these two elements.  Now, following the arrows around the diagram, we see that $r^2$ must pair with $i^2=-1$ and $r^3$ corresponds to $i^3=-i$.  In summary, the isomorphism between $R_4$ and $\langle i\rangle$ (in $Q_8$) is given by $e\leftrightarrow 1$, $r\leftrightarrow i$, $r^2\leftrightarrow -1$, and $r^3\leftrightarrow -i$, which is depicted in Figure~\ref{fig:isoR4}. Note that this time we have not recolored the edges so that they match.  Nonetheless, the correspondence should be clear.
%
%\tikzstyle{vert} = [circle, draw, fill=grey,inner sep=0pt, minimum size=6.5mm]
%
%\begin{figure}[!ht]
%\centering
%\begin{tikzpicture}[scale=1.5,auto]
%\node (r) at (1,1) [vert] {\scriptsize $r$};
%\node (e) at (-1,1) [vert] {\scriptsize $e$};
%\node (r3) at (-1,-1) [vert] {\scriptsize $r^3$};
%\node (r2) at (1,-1) [vert] {\scriptsize $r^2$};
%
%\path[r] (e) to (r);
%\path[r] (r) to (r2);
%\path[r] (r2) to (r3);
%\path[r] (r3) to (e);
%
%\node (R4) at (0,0) {$R_4$};
%
%\node (i) at (5,1) [vert] {\scriptsize $i$};
%\node (1) at (3,1) [vert] {\scriptsize $1$};
%\node (-i) at (3,-1) [vert] {\scriptsize $-i$};
%\node (-1) at (5,-1) [vert] {\scriptsize $-1$};
%
%\path[b] (1) to (i);
%\path[b] (i) to (-1);
%\path[b] (-1) to (-i);
%\path[b] (-i) to (1);
%
%\node (subgroupQ8) at (4,0) {$\langle i\rangle\leq Q_8$};
%
%\path[correspondence] (e) to [bend left=30] (1);
%\path[correspondence] (r) to [bend left=30] (i);
%\path[correspondence] (r2) to [bend right=30] (-1);
%\path[correspondence] (r3) to [bend right=30] (-i);
%
%\end{tikzpicture}
%\caption{Isomorphism between $\langle i\rangle\leq Q_8$ and $R_4$.}\label{fig:isoR4}
%\end{figure}
%
%Now, take a look at the Cayley diagram for $D_4$ with generating set $\{r,s\}$.  As we noticed in Exercise~\ref{exer:R4_in_D4}, $R_4$ is a subgroup of $D_4$.  We could say that this subgroup is isomorphic to $R_4$, but in this case, we can say something even stronger: they are equal!
%
%Before continuing, we need to emphasize an important point.  If the Cayley diagram for one group does not look like the Cayley diagram for another group, then that does \emph{not} immediately imply that the groups are not isomorphic.  The issue is that perhaps we could choose appropriate generating sets for each group so that the Cayley diagrams do look alike.  For example, notice that our standard Cayley diagram for $R_4$ does not look like the Cayley diagram that you constructed for $V_4$ in Exercise~\ref{exer:V4}.  This does \emph{not} imply that these two groups are not isomorphic.  We would need to do some more work in order to determine whether the two groups are isomorphic or not.  You will get a chance to do this in Exercise~\ref{exer:R4vsV4}.
%
%It turns out that there is a fancy word for the size of a group.
%
%\begin{definition}
%If $G$ is a group with $n$ distinct actions, then we say that $G$ has \textbf{order} $n$ and write $|G|=n$.  If $G$ contains infinitely many elements, then we say $G$ has infinite order and write $|G|=\infty$.
%\end{definition}
%
%\begin{exercise}
%Find the orders of the following groups: $S_2$, $\Spin_{1\times 2}$, $\Spin_{3\times 3}$, $R_4$, $D_3$, $D_4$, $V_4$, and $Q_8$.
%\end{exercise}
%
%\begin{theorem}\label{thm:iso_same_order}
%Suppose $G$ and $G'$ are two groups of actions such that $G\cong G'$.  Then $|G|=|G'|$.
%\end{theorem}
%
%Unfortunately, the converse of the previous theorem is not true in general.  That is, two groups that have the same order may or may not be isomorphic.
%
%Loosely speaking, if one group has a property that the other does not have, then the two groups cannot be isomorphic.  For example, if one group has the property that every pair of actions commutes (i.e., the order\footnote{Don't confuse the word ``order" in this sentence with the order of a group.} of the actions does not matter), but another group has a pair of actions that do not commute, then the two groups cannot be isomorphic.  Moreover, if one group contains an action that requires a minimum of $k$ applications to get back to the do-nothing action, but a second group does not have such an element, then the two groups cannot be isomorphic.  
%
%Justifying these two claims takes a bit of work and for now, we'll put that on hold.  For the time being, if you don't see why these claims about when two groups are not isomorphic are true, just take them on faith and we will return to the issue in a later chapter.  Feel free to use these ideas in the exercises that follow.
%
%\begin{problem}\label{exer:R4vsV4}
%Determine whether $R_4$ and $V_4$ are isomorphic.  Justify your answer.  If they are isomorphic, specify the isomorphism by listing the correspondence of elements.  If they are not isomorphic, explain why.
%\end{problem}
%
%\begin{problem}\label{prob:R6_not_iso_D3}
%Consider the group given by the Cayley diagram for $R_6$ that was given in Exercise~\ref{exer:intro_R6}.  We can think of $R_6$ as the rotation group for a regular hexagon.  Determine whether $R_6$ and $D_3$ are isomorphic.  Justify your answer.  If they are isomorphic, specify the isomorphism by listing the correspondence of elements.  If they are not isomorphic, explain why.
%\end{problem}
%
%\begin{exercise}
%Consider two light switches on a wall side by side.  Consider the group of actions that consists of all possible actions that you can do to the two light switches.  For example, one action is toggle the left light switch while leaving the right alone.  Let's call this group $L_2$.
%\begin{enumerate}[label=\rm{(\alph*)}]
%\item How many distinct actions does $L_2$ have?
%\item Can you find a minimal generating set for $L_2$?  If so, give these actions names and then write all of the actions of $L_2$ as words in your generator(s).
%\item Using your generators from part (b), draw a Cayley diagram for $L_2$.
%\end{enumerate}
%\end{exercise}
%
%\begin{problem}
%Determine whether $L_2$ and $V_4$ are isomorphic.  Justify your answer.  If they are isomorphic, specify the isomorphism by listing the correspondence of elements. If they are not isomorphic, explain why.
%\end{problem}
%
%\begin{problem}
%Determine whether $Q_8$ and $D_4$ are isomorphic.  Justify your answer.  If they are isomorphic, specify the isomorphism by listing the correspondence of elements. If they are not isomorphic, explain why.
%\end{problem}
%
%\begin{problem}
%Determine whether $\Spin_{1\times 2}$ and $D_4$ are isomorphic.  Justify your answer.  If they are isomorphic, specify the isomorphism by listing the correspondence of elements. If they are not isomorphic, explain why.
%\end{problem}
%
%\begin{exercise}\label{exer:S3}
%Consider the group that acts on three coins that are in a row by rearranging their positions (but not flipping them over).  This group is called $S_3$.  Number the positions of the coins (not the coins themselves) 1, 2, 3 from left to right.  Let $s_1$ be the action that swaps the coins in positions 1 and 2 and let $s_2$ be the action that swaps the coins in positions 2 and 3.
%\begin{enumerate}[label=\rm{(\alph*)}]
%\item The group $S_3$ consists of 6 actions, which we can generate with $s_1$ and $s_2$.  Write all 6 actions as words in $s_1$ and $s_2$.
%\item Using $s_1$ and $s_2$ as generators, draw a Cayley diagram for $S_3$.
%\end{enumerate}
%\end{exercise}
%
%\begin{problem}\label{prob:D3_iso_S3}
%Determine whether $S_3$ and $D_3$ are isomorphic.  Justify your answer.  If they are isomorphic, specify the isomorphism by listing the correspondence of elements.  If they are not isomorphic, explain why. Don't forget that we've drawn two different Cayley diagrams for $D_3$.
%\end{problem}


% We already know that $R_4$ and $V_4$ are not isomorphic, but this becomes even more apparent if you compare their subgroup lattices.

%\begin{problem}
%What claims can be made about the subgroup lattices of two groups that are isomorphic? What claims can be made about the subgroup lattices of two groups that are not isomorphic?  What claims can be made about two groups if their subgroup lattices look nothing alike?  \emph{Hint:} The answers to two of these questions should be obvious, but the answer to the remaining question should be something like, ``we don't have enough information to make any claims."
%\end{problem}



%Consider three light switches on a wall side by side.  Consider the group of actions that consists of all possible actions that you can do to the three light switches (involving ``on" and ``off").  Let's call this group $L_3$. It should be easy to see that $L_3$ has 8 distinct actions.
%\begin{enumerate}[label=\rm{(\alph*)}]
%\item Can you find a minimal generating set for $L_3$?  If so, give these actions names and then write all of the actions of $L_3$ as words in your generator(s).
%\item Using your generating set from part (a), draw a Cayley diagram for $L_3$.
%\item Is $L_3$ cyclic? Briefly justify your answer.
%\item Is $L_3$ abelian? Briefly justify your answer.
%\end{enumerate}


%scraps:
%Here are two final problems to conclude this section.
%
%\begin{problem}
%Several times we've referred to the fact that some subgroups are visible in a Cayley diagram for the parent group and some subgroups are not.  Suppose $(G,*)$ is a group and let $H\leq G$.  Can you describe a process for creating a Cayley diagram for $G$ that ``reveals" the subgroup $H$ inside of this Cayley diagram?
%\end{problem}
%
%\begin{problem}
%Suppose $(G,*)$ is a finite group and let $H\leq G$.  Can you describe a process that ``reveals" the subgroup $H$ inside the group table for $G$?  Where will the clones for $H$ end up?
%\end{problem}



\end{section}
